Feature Selection for Mixed-Effects Models

### 2022-07-04 presentation/general_presentation.tex
Fill presentation up to algorithm's pseudocode

\usepackage{xspace}
\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
\newcommand{\ouralgo}{R\&S-Mixed }
\title{Feature Selection for Mixed-Effects Models}
\date{Thursday, 3\textsuperscript{rd} of December, 2020}
\author{Aleksei Sholokhov}
% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}

@@

\section{Introduction}
\subsection{Linear Mixed-Effects Models}
\begin{frame}{Linear Mixed-Effect Models are a type of regression models for grouped data}

	\[
	\text{Dataset: m groups } (X_i, y_i),\quad i = 1, \dots m
    \]
	\begin{columns}[T,onlytextwidth]
    \column{0.5\textwidth}
	 \centering Standard Linear Regression:
   	\begin{figure}
   		\includegraphics[width=0.9\textwidth]{Figures/lme_example_mean_prediction}
   	\end{figure}
   	\[
   		y = X\beta + \varepsilon, \quad \varepsilon \sim \NN(0, \Lambda) 
   	\]

    \column{0.5\textwidth}
    	\centering  Linear Mixed-Effect Model:
   	\begin{figure}
   		\includegraphics[width=0.9\textwidth]{Figures/lme_example_random_prediction}
   	\end{figure}
   	   		\[
   		\begin{split}
   			y_i & = X_i(\beta + {\color{red} u_i}) + \varepsilon_i, \quad \varepsilon_i \sim \NN(0, \Lambda_i) \\
   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma)}
   		\end{split}
   		\]

@@

\begin{frame}{Notation}
	\eq{
   		y_i & = \alt<3->{X_i{\color<4,9>{red}\beta} + {\color<3>{red}Z_i}{\color<5,9>{red}u_i}}{{\color<2>{red}X_i}(\beta + u_i)} + \varepsilon_i \quad i = 1\dots m \\
   		{\color<7>{red}\varepsilon_i} & \sim \NN(0, {\color<8,9>{red}\Lambda_i}) \\
   		u_i & \sim \NN(0, {\color<6,9>{red}\Gamma})
   	}

   	\begin{itemize}
   		\item<2-> $X_i \in \R^{n_i \times p}$ -- \only<3>{{\color{red} fixed}}\only<4->{fixed} features, or covariates, \only<3>{{\color{red} $Z_i \in \R^{n_i \times q}$ -- random features }}\only<4->{ $Z_i \in R^{n_i \times q}$ -- random features}
   		\item<3-> $p$ -- number of fixed features, $q$ -- number of random effects.
   		\item<4-> $\beta \in \R^p$ -- fixed effects, or mean effects
   		\item<5-> $u_i \in \R^q$ -- random effects
   		\item<6-> $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects
   		\item<7-> $\varepsilon_i \in \R^{n_i}$ -- observation noise
   		\item<8-> $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for noise
   		\item<9-> Unknowns: $\beta$, $u_i$, $\Gamma$, sometimes $\Lambda_i$.

   	\end{itemize}
   	\uncover<10->{
   	\[
   		\Gamma = \diag{({\color{red}\gamma})},\ {\color{red}\gamma \in \R^{q}} \quad \text{ and } \Lambda_i \text{ are known.}
   	\]
   	}
\end{frame}

\begin{frame}[fragile]{Likelihood for Mixed Models}
Suppose we have a linear mixed effects model

\eq{
	\label{eq:lmm_setup}
	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i \\
	u_i & \sim \NN(0, \Gamma), \ \Gamma = \diag{\gamma} \\
	\varepsilon_i & \sim \NN(0, \Lambda_i)\\
}
where $\beta$, $\gamma$, and $u_i$ are unknown model parameters, and $\lambda_i$ are known.

	Noticing that $Z_iu_i + \varepsilon_i \sim \NN(0, Z_i\Gamma Z_i^T + \Lambda_i)$ we get the negative log-likelihood:
	 \eq{
	\label{eq:lmm_objective}
	\mathcal{L}(\beta, \gamma) & = \sum_{i = 1}^m \half(y_i - X_i\beta)^T(Z_i\Gamma Z_i^T + \Lambda_i)^{-1}(y_i - X_i\beta) + \\ & + \half\log{\det{\pa{Z_i \Gamma Z_i^T + \Lambda_i}}}
	}

@@

\subsection{Feature Selection for Mixed Models}
\begin{frame}{Feature Selection for Mixed Models}

Consider the following feature selection setup:

\eq{
	\min_{\beta, \gamma} & \LL(\beta, \gamma) \\
	\text{s.t. } & \|\beta\|_0 \leq k \\
	& \|\gamma\|_0 \leq j \\
	& \gamma \geq 0
}

where $k$ and $j$ is the size of the sparse support for the fixed and random effects respectively.

This is a very hard problem to solve directly.

@@

\begin{frame}{Methods for Feature Selection}
There are several ways to address this setup:
  \begin{columns}[T,onlytextwidth]
    \column{0.33\textwidth}
    (0) Exhaustive Search:
\[
	\begin{split}
	\min_{\beta, \gamma} & \LL(\beta, \gamma) \\
	\text{s.t. } & \|\beta\|_0 \leq k \\
	& \|\gamma\|_0 \leq j \\
	& \gamma \geq 0
	\end{split}
\]

    \column{0.33\textwidth}
    \uncover<2->{
    \centering
    (1) $\ell_1$-relaxation:
\[
	\begin{split}
	\min_{\beta, \gamma} & \LL(\beta, \gamma) \\
	\text{s.t. } & \|\beta\|_{\color{red}1} \leq {\color{red} \lambda_\beta} \\
	& \|\gamma\|_{\color{red}1} \leq {\color{red} \lambda_\gamma} \\
	& \gamma \geq 0
	\end{split}
\]
	}
	\column{0.33\textwidth}
	\uncover<3->{
    \centering
    (3) Other types of sparsity-inducing penalties\\ ($\ell_p$, $SCAD$).
    }
  \end{columns}

 \onslide<1-3>{
	\begin{columns}[T, onlytextwidth]
	\column{0.33\textwidth}
	\onslide<1-3>{
	\centering
	AIC \cite{Vaida2005}\\
	BIC \cite{Jones2011}\\
	}
	\column{0.33\textwidth}
	\onslide<2-3>{
	\centering
	M-ALASSO \cite{Bondell2010} \\
	ALASSO \cite{Lin2013} \\
	}
	\column{0.33\textwidth}
	\onslide<3>{
	\centering
	SCAD \cite{Fan2001}\\
	SCAD-P \cite{Fan2012}
	}
	\end{columns}
}

\onslide<4->{
	\begin{center}
	(2) SR3-type Relaxation \cite{Zheng2019}
	\eq{
	\min_{\beta, \gamma, {\color{red}\hbeta}, {\color{red}\tgamma}} & \LL(\beta, \gamma) + {\color{red}\frac{\lambda_\beta}{2}\|\beta - \tbeta\|^2_2 + \frac{\lambda_{\gamma}}{2}\|\gamma - \tgamma\|_2^2}\\
	\text{s.t. } 	& \|{\color{red}\tbeta}\|_0 \leq k \\
					& \|{\color{red}\tgamma}\|_0 \leq j, \  \gamma \geq 0
	}
	\end{center}
	}

\end{frame}
\section{Proposed Algorithm}
\begin{frame}{Relax-and-Split from Optimization Perspective}
\eq{
	{\color<2, 3>{red}\min}_{{\color<2>{red}\beta, \gamma,} {\color<3>{red}\tbeta, \tgamma}} & {\color<2>{red}\LL(\beta, \gamma)} + {\color<2,3>{red}\frac{{\color<4>{red}\lambda_\beta}}{2}\|\beta - \tbeta\|^2_2} + {\color<2,3>{red}\frac{{\color<4>{red}\lambda_\gamma}}{2}\|\gamma - \tgamma\|_2^2}\\
	\text{s.t. } 	& {\color<3>{red}\|\tbeta\|_0 \leq k, \ \|\tgamma\|_0 \leq j} \\ 
		& {\color<2>{red}\gamma \geq 0}, \ {\color<3>{red}\tgamma \geq 0}
	}
We'll find $\beta^*$, $\gamma^*$, $\tbeta^*$, and $\tgamma^*$ via alternating partial minimization:

\begin{itemize}
	\item<2-> Minimization w.r.t. $\beta$ and $\gamma$ is done via \textbf{Interior Point Method}.
	\item<3-> Minimization w.r.t. $\tbeta$ and $\tgamma$ is \textbf{projection} of $\beta$ and $\gamma$ onto $\ell_0$-balls.
	\item<4-> $\lambda_\beta$ and $\lambda_\gamma$ are \textbf{iteratively increased} until $\tbeta \approx \beta$ and $\tgamma \approx \gamma$.
\end{itemize}
\only<-4>{
\visible<3>{
	\eq{
		\tbeta^{(k+1)} \leftarrow \proj_{\|\beta\|_0 \leq k}(\beta^{(k)}) \quad &\implies \quad \text{Take max k from } \beta \text{, rest to 0} \\
		\tgamma^{(k+1)} \leftarrow \proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)})\quad &\implies \quad \text{Take max j from }\gamma \text{, rest to 0}
		}
}
}
\only<5->{
\textbf{Advantages}:
\begin{enumerate}
	\item Interpretable hyperparameters.
	\item No model bias when the $k-$subspace is chosen.
\end{enumerate}
}
\end{frame}

%\begin{frame}{Interior Point Method}
%The Lagrangian of the barrier relaxation is:
%	\eq{
%	F_\mu(v, \beta, \gamma) & = \LL(\beta, \gamma) +  \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2 + \\ & + \mu\sum_{i=1}^{q}\log(\gamma_i) - v^T\gamma \\
%	}
%	The descent direction is chosen as:
%	\eq{
%	\nabla^2F_\mu\begin{bmatrix}
%		\Delta v \\
%		\Delta \beta \\
%		\Delta \gamma
%	\end{bmatrix} = -\nabla F_\mu\begin{bmatrix}
%		\Delta v \\
%		\Delta \beta \\
%		\Delta \gamma
%	\end{bmatrix}
%	\text{, step is: }
%	\begin{cases}
%	v^{(k+1)} = v^{(k)} + \alpha_k \Delta v \\
%	\gamma^{(k+1)} = \gamma^{(k)} + \alpha_k \Delta\gamma \\
%	\beta^{(k+1)} = \beta^{(k)} + \alpha_k \Delta \beta
%	\end{cases}
%	}
%	and $\alpha_k$ is chosen such that
%	\eq{
%	& \gamma^{(k+1)} \geq 0,\ v^{(k+1)} \geq 0 \\
%	& \|\nabla F_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|\nabla F_\mu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\| 
%	}
%\end{frame}

\begin{frame}{R\&S-Mixed}
The pseudocode of the proposed algorithm:
{\small
\begin{algorithm}[H]
	$\lambda_\beta = 0$; $\lambda_\gamma = 0$ \\
	\Repeat{$\tbeta \approx \beta, \tgamma \approx \gamma$}{
	$\lambda_\beta \leftarrow 2(1 + \lambda_\beta)$ \\
	$\lambda_\gamma \leftarrow 2(1 + \lambda_\gamma)$\\
	\Repeat{converges}{
		$\tbeta^{(k+1)} \leftarrow \proj_{\|\beta\|_0 \leq k}(\beta^{(k)})$\\ 
		$\tgamma^{(k+1)} \leftarrow \proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)})$ \\
		$\beta^{(k+1)}, \gamma^{(k+1)} \leftarrow \argmin_{\gamma \geq 0, \beta}\LL(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2$
	}
	}
	\BlankLine
\end{algorithm}
	}
\end{frame}

>>>>

\usepackage{xspace}
\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
\newcommand{\ouralgo}{\ensuremath{\mathcal{MSR}3}}
\title{Feature Selection for Mixed-Effects Models}
\date{\today}
\author{Aleksei Sholokhov}
% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}

@@

\section{Feature Selection for Linear Mixed-Effect Models}
\subsection{Linear Mixed-Effects Models}
\begin{frame}{Linear Mixed-Effect Models}

Linear Mixed-Effect (LME) models are often used for analyzing combined data across a range of groups. \newline

They use use covariates to separate the population variability (fixed-effects) from the group variability (random effects). \newline

LMEs borrow strength across groups to estimate key statistics in cases when data within units are sparse or highly variable. \newline

\begin{frame}{Linear Mixed-Effect Models}

	Dataset: $m$ groups $(X_i, Z_i, y_i),\quad i = 1, \dots m$, each has $n_i$ observations
	\begin{itemize}
		\item 	$X_i \in \R^{n_i \times p}$ -- group $i$ design matrix for fixed features
		\item 	$Z_i \in \R^{n_i \times q}$ -- group $i$ design matrix for random effects
		\item 	$y_i \in \R^{n_i}$ -- group $i$ observations  
	\end{itemize}

	\begin{columns}[T,onlytextwidth]
    \column{0.5\textwidth}
	 \centering Standard Linear Regression:
   	\begin{figure}
   		\includegraphics[width=0.9\textwidth]{Figures/lme_example_mean_prediction}
   	\end{figure}
   	\[
   		y = X\beta + \varepsilon, \quad \varepsilon \sim \NN(0, \Lambda) 
   	\]

    \column{0.5\textwidth}
    	\centering  Linear Mixed-Effect Model:
   	\begin{figure}
   		\includegraphics[width=0.9\textwidth]{Figures/lme_example_random_prediction}
   	\end{figure}
   	   		\[
   		\begin{split}
   			y_i & = X_i\beta + {\color{red}Z_i u_i} + \varepsilon_i, \quad \varepsilon_i \sim \NN(0, \Lambda_i) \\
   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma4)}
   		\end{split}
   		\]

@@

\begin{frame}{Notation}
	\eq{
   		y_i & = X_i\beta + Z_iu_i + \varepsilon_i \quad i = 1\dots m \\
   		\varepsilon_i & \sim \NN(0, \Lambda_i) \\
   		u_i & \sim \NN(0, \Gamma)
   	}

   	\begin{itemize}
   		\item $p$ -- number of fixed features, $q$ -- number of random effects.
   		\item $\beta \in \R^p$ -- fixed effects, or mean effects
   		\item $u_i \in \R^q$ -- random effects
   		\item $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects, often $\Gamma = \diag{(\gamma)}$
   		\item $\varepsilon_i \in \R^{n_i}$ -- observation noise
   		\item $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for noise
   	\end{itemize}
   	Unknowns: $\beta$, $u_i$, $\gamma$, sometimes $\Lambda_i$.
\end{frame}

\begin{frame}{Likelihood for Mixed Models}
Negative log-likelihood:
\eq{
	\label{eq:lmm_objective}
	\mathcal{L}(\beta, \gamma) & = \sum_{i = 1}^m \half(y_i - X_i\beta)^T(Z_i\Gamma Z_i^T + \Lambda_i)^{-1}(y_i - X_i\beta) + \\ & + \half\log{\det{\pa{Z_i \Gamma Z_i^T + \Lambda_i}}}, \quad \Gamma = \diag{(\gamma)}
	}
Maximum likelihood estimates for $\beta$ and $\gamma$ solve the problem:
\eq{
	\mathcal{LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma)
}

To select covariates we add a sparsity-promoting regularizer $R(\beta, \gamma)$
\eq{
	\mathcal{FS-LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma) + R(\beta, \gamma)
}

\begin{itemize}
	\item $\LL(\beta, \gamma)$ is smooth on its domain, quadratic w.r.t. $\beta$ and $\bar\eta$-weakly-convex w.r.t. $\gamma$.
	\item $R(\beta, \gamma)$ is closed, proper, convex, with easily computed \textit{prox operator}
\end{itemize}

@@

\begin{frame}{Regularization}
\begin{itemize}
	\item $R(\beta, \gamma)$ is closed, proper, convex, with easily computed \textit{prox operator}
\end{itemize}

\eq{
	\prox_{\alpha R + \delta_{\CC}}(\tbeta, \tgamma) & := \argmin_{(\beta, \gamma) \in \CC} R(\beta, \gamma) + \frac{1}{2\alpha}\|(\beta, \gamma) - (\tbeta, \tgamma)\|_2^2, \\ & \text{ where } \CC := \R^p \times R^q_+ \\
}

Examples: 
\begin{itemize}
	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \cite{Bondell2010,Lin2013}
	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \cite{Vaida2005,Jones2011}
	\item $R(x)$ -- SCAD penalty (\cite{Fan2001,Fan2012})
\end{itemize}

TODO: add picture

@@

\begin{frame}{SR3-Relaxation for Mixed-Effect Models}
Original problem $\mathcal{FS-LME}$:
\eq{
	\min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma) + R(\beta, \gamma)
}
Relaxed problem $\ouralgo$:
\eq{
	\label{eq:msr3_formulation_explicit}
	\min_{\beta, \tbeta \in \R^p,\, \gamma, \tgamma \in \R^{q}_+} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma) + R(\beta, \gamma)
}
where the \textit{relaxation} $\kappa_\eta$ decouples the likelihood and the regularizer 
\eq{
	\kappa_\eta(\beta - \tbeta, \gamma - \tgamma) := \frac{\eta}{2}\|\beta - \tbeta\|^2_2 + \frac{\eta + \bar\eta}{2}\|\gamma - \tgamma\|_2^2
}
and the \textit{projection function} $\phi_\mu$ replaces $\gamma \geq 0$ with a log-barrier 
\eq{
	\phi_\mu(\gamma) := \begin{cases}
		-\mu\sum_{i=1}^q \ln(\gamma_i/\mu), & \mu > 0 \\
		\delta_{\R_+^q}(\gamma), & \mu = 0 \\
		+\infty, & \mu < 0
	\end{cases}
} 
\end{frame}

\begin{frame}{Value Function Reformulation}
$\ouralgo$-relaxation replaces the original likelihood $\LL$ with a \textit{value function} $u_{\eta,\mu}$:
\eq{
	\label{eq:value_function_definition}
	u_{\eta,\mu}(\tbeta,\tgamma) & := \min_{(\beta, \gamma)} \LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\  & := \min_{(\beta, \gamma)} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma)
}
so $\ouralgo$-formulation~(\ref{eq:msr3_formulation_explicit}) becomes
\eq{
	\min_{(\tbeta, \tgamma) \in \CC} u_{\eta,\mu}(\tbeta, \tgamma) + R(\tbeta, \tgamma)
}
When $\bar\eta$ is larger than the weak-convexity constant
\begin{itemize}
	\item $u_{\eta,\mu}$ is well-defined and continuously differentiable.
	\item Solutions $(\tbeta^*, \tgamma^*)$ for $\ouralgo$ converge to solutions $(\beta^*, \gamma^*)$ of $\mathcal{FS-LME}$ \\ when $\mu \rightarrow 0$ and $\eta \rightarrow \infty$.
\end{itemize}

\textbf{Key observation}: in practice, we don't need accurate solutions for~(\ref{eq:value_function_definition}): a few Newton iterations is enough.
\end{frame}

\begin{frame}{Value Function Reformulation}
TODO: insert image here
\end{frame}

\begin{frame}{Designing an Algorithm}
	Gradient of a Lagrangian:
	\eq{
		G_{\nu,\eta}((\beta,\gamma,v),(\tbeta,\tgamma)) := \begin{bmatrix}
			\nabla_\beta \LL(\beta, \gamma) + \eta(\beta-\tbeta) \\
			\nabla_\gamma \LL(\beta, \gamma) + (\bar\eta + \eta)(\gamma-\tgamma) - v\\
			v \bigodot \gamma - \mu\textbf{1}
		\end{bmatrix}
	}

\textbf{Lemma:} For every $(\mu,\eta) \in \R_+\times\R_{++}$,
\eq{
	(\hat\beta,\hat\gamma) = \argmin_{(\beta,\gamma)}\LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\
	\text{(equivalent to)} \\
	\exists \hat{v} \in \R_{+}^q \text{ s.t. } G_{\nu,\eta}((\beta,\gamma,\hat{v}),(\tbeta,\tgamma)) = 0
}
If $\mu > 0$, then $\hat{v} = -\nabla\phi_\mu(\hat{\gamma})$, and if $\mu = 0$, then $\hat{v}$ is the unique KKT multiplier associated with the constraint $0 \leq \gamma$.
\end{frame}

\begin{frame}{$\ouralgo$ Algorithm}
\input{../extras/08Pseudocode}
\end{frame}
